### 2021-09-23 revision/ch1intro.tex
uh

Landscapes evolve when water, wind, and ice, under the influence of gravity and sunlight, erode along topographic gradients produced by tectonics.
Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands.
Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.

@@

Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}. The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018, Masteller2019, Pretzlav2020}.
Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values \citep{Ancey2020b}.

@@

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport theory is to predict the downstream movement of individual particles.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow, but precise relationships between the fluid flow and the applied forces remain unknown \citep{Maxey1983,Schmeeckle2007}.
Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.

@@

Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}. 
These buried particles cannot move along the bed surface again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of relative immobility \citep{Hassan1994,Ferguson2002}.

@@

\citet{Nikora2001,Nikora2002} provided a conceptual framework which helps to organize this complexity.
Nikora et al divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
The intermediate range reflects particle motions through sequences of bed encounters, when particles alternately accelerate and decelerate. 
The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017}.
This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize the literature describing trajectories of individual grains.

\subsection{Motivation: tracers and basic understanding}
An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate virtual velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}. This obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.

@@

Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
As his interest was on the global range of particle motion when many motion-rest alternations have occurred, and because the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
With this approximation, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.

Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed. In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this was not formalized until much later \citep{Montroll1965}.

Einstein's assumptions imply that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
\be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
where $\dot{x} = dx/dt$, and $\mu(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
A particular realization of this pulsed noise can be written
\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i). \label{eq:einrando} \ee
Here $N(t)$ is the number of particle entrainments in time $t$ which is  a Poisson random variable, distributed as $P(N) = e^{-k_E t} (k_E t)^N/N!$.
The $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
Figure \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as a sequence of steps and rests.

Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (equation \ref{eq:einrando}).
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
	\caption{Panel (a) indicates the representation of Einstein's model as an idealized ``white shot noise", as indicated in eq. \ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$). }
	\label{fig:einsteinfig}
\end{figure}

@@

Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales. They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
\be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
%where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
This noise is displayed in figure \ref{fig:lislefig} panel (a). Some particle trajectories produced by equation \ref{eq:lislelangevin} are displayed in figure \ref{fig:lislefig} panel (b).

@@

For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of equation \ref{eq:lislemaster} is \citep{Lisle1998}
\begin{multline}  P(x,t) =  e^{-\chi-\tau} \Big[ \frac{k_E}{V}\delta(\tau) + \frac{k_E}{V} \sqrt{\frac{\chi}{\tau}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_D}{V}\mathcal{I}_0(2\sqrt{\chi\tau}) \\
	+ \frac{k_Ek_D}{kV}\sqrt{\frac{\tau}{\chi}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_Ek_D}{kV} \delta(\chi)
	\Big]\theta(\chi)\theta(\tau), \label{eq:lisledist}
\end{multline}
where $\chi = k_D x/V$ and $\tau = k_E(t-x/V)$ are shorthand notations.
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with an effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state. In contrast, the variance develops a two range scaling.
Computing $\sigma_x^2$ provides
\be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) , \label{eq:lislevar}\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$. 
The rapid spreading of particles at short timescales occurs because some are completely stationary while others move with velocity $V$.

@@

Some authors have modelled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically. Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.

@@

The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow. Particle-particle interactions are represented with the discrete element method, where particles are represented as spheres \citep{Cudnall1979} or polytopes \citep{Wachs2019} while forces are phrased as idealized spring-dashpot models.
The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
Particles are represented either as spheres coupled to the flow \citep{}, 
 while particles provide the boundary condition for the flow \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.

@@

\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities ($u$) with the Langevin equation
\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by equation \ref{eq:fanlangevin} is
\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
This is the (two-sided) exponential distribution observed in one subset of experiments.

\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance. The timescale $t_r$ is a relaxation time over which velocity fluctuations decay. This time, the Fokker-Planck equation is
\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
and the steady state solution becomes
\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
which is the Gaussian velocity distribution from the other subset of the experiments.
These two models of bedload velocity distributions are summarized in figure \ref{fig:fanAncey}.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag. Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d). The latter has relatively narrow and symmetric fluctuations, with a positive mode well above $u=0$.}
	\label{fig:fanAncey}
\end{figure}
Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and other distributions have been reported besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
A comprehensive model of the full range of bedload transport observations is a desirable target.

\subsection{The definition of the flux}
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.pdf}
	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$. Here, particles are shown moving from time $t-dt$ (faint white particles) to time $t$ (grey particles). Each particle's velocity component $u_i$ in the downstream direction is indicated. At the instant $t$, the surface definition of the flux \ref{eq:surfflux} gives $q = \frac{1}{L}(u_7+u_8).$, while the volume definition \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_6)$,
	showing the non-equivalence of these two definitions.}
	\label{fig:fluxdefs}
\end{figure}
Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
Today, there are two main complementary definitions of the bedload flux.

The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
\be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.

@@

One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}. This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$. These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}, in what can be characterized as a mixture of the Einstein and Bagnold strategies.

>>>>

Landscapes evolve when water, wind, and ice erode along the gradients produced by tectonics.
Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands.
Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.

@@

Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulence characteristics \citep{Singh2010, Santos2014, Liu2016}.
The surface arrangement and consolidation of grains affects sediment mobility \citep{Miller1966,Paintal1971,Dwivedi2012}, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018, Masteller2019, Pretzlav2020}.
Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values \citep{Ancey2020b}.

@@

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes, highlight the shortcomings to be addressed, and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport theory is to predict the downstream movement of individual particles.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow, but relationships between the fluid flow and the applied forces are known only approximately \citep{Michaelides1997a,Schmeeckle2007}.
Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.

@@

Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}. 
These buried particles cannot move appreciably until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of relative immobility \citep{Hassan1994,Ferguson2002}.

@@

\citet{Nikora2001,Nikora2002} provided a conceptual framework which helps to organize this complexity.
Nikora et al. divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
The intermediate range reflects particle motions through sequences of bed encounters, when particles alternately accelerate and decelerate. 
The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017}.
This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize the literature describing trajectories of individual grains.

\subsection{Motivation: tracers and basic understanding}

An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate virtual velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}. This obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.

@@

Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
As his interest was on the global range of particle motion when many motion-rest alternations have occurred, and because the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
With this approximation, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length and rests of random duration. 
Implicitly, this concept of sediment trajectories assumes that movement velocities are infinite.

Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed. In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this approach not formalized until much later \citep{Montroll1965}. In Earth science, random walks have been widely applied to hydrologic problems \citep{Berkowitz2006}, but their application to surface processes remains relatively uncommon \citep[e.g.][]{Schumer2009}.

Einstein's assumptions imply that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
\be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
where $\dot{x} = dx/dt$, and $\mu(t)$ is a Poisson pulse noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
A particular realization of this pulsed noise can be written
\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i). \label{eq:einrando} \ee
Here $N(t)$ is the number of particle entrainments in time $t$ which is  a Poisson random variable, distributed as $P(N) = e^{-k_E t} (k_E t)^N/N!$.
The $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
Figure \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as a sequence of steps and rests.

Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (eq. \ref{eq:einrando}).
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
	\caption{Panel (a) indicates the representation of Einstein's model as an idealized Poisson pulse noise, as indicated in eq. \ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$). }
	\label{fig:einsteinfig}
\end{figure}

@@

Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales. They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
\be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
%where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
This noise is displayed in fig. \ref{fig:lislefig} panel (a). Some particle trajectories produced by eq. \ref{eq:lislelangevin} are displayed in fig. \ref{fig:lislefig} panel (b).

@@

For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of eq. \ref{eq:lislemaster} is \citep{Lisle1998}
\begin{multline}  P(x,t) =  e^{-\chi-\tau} \Big[ \frac{k_E}{V}\delta(\tau) + \frac{k_E}{V} \sqrt{\frac{\chi}{\tau}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_D}{V}\mathcal{I}_0(2\sqrt{\chi\tau}) \\
	+ \frac{k_Ek_D}{kV}\sqrt{\frac{\tau}{\chi}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_Ek_D}{kV} \delta(\chi)
	\Big]\theta(\chi)\theta(\tau), \label{eq:lisledist}
\end{multline}
where $\chi = k_D x/V$ and $\tau = k_E(t-x/V)$ are shorthand notations.
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with an effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state. In contrast, the variance develops a two range scaling.
Computing $\sigma_x^2$ provides
\be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) , \label{eq:lislevar}\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, eq. \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$. 
The rapid spreading of particles at short timescales occurs because some are completely stationary while others move with velocity $V$.

@@

Some authors have modeled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically. Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.

@@

The state of the art within this category of sediment transport models is to simulate both fluid and particle phases simultaneously.
Particle-particle interactions are represented with the discrete element method, where particles are represented as spheres \citep{Cudnall1979} or polytopes \citep{Wachs2019} while forces are phrased as idealized spring-dashpot models.
The fluid flow is modeled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
A majority of these approaches simulate the fluid without including no-slip boundary conditions on the particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although works are beginning to include the fluid boundary conditions on individual particles with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.

@@

\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities ($u$) with the Langevin equation
\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from eq. \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by eq. \ref{eq:fanlangevin} is
\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
This is the (two-sided) exponential distribution observed in one subset of experiments.

\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance. The timescale $t_r$ is a relaxation time over which velocity fluctuations decay. This time, the Fokker-Planck equation is
\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
and the steady state solution becomes
\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
which is the Gaussian velocity distribution from the other subset of the experiments.
These two models of bedload velocity distributions are summarized in fig. \ref{fig:fanAncey}.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag. Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d). The latter has relatively narrow and symmetric fluctuations, with a positive mode well above $u=0$.}
	\label{fig:fanAncey}
\end{figure}
Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and other distributions have been reported besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
A comprehensive model of the full range of bedload transport observations remains a desirable target.

\subsection{The definition of the flux}
% for some reason the pdf of this figure does not properly render 
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.png}
	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$. Here, particles are shown moving from time $t-\Delta t$ (faint white particles) to time $t$ (grey particles). Each particle's downstream displacement $\Delta x_i$ is indicated. These displacements define the downstream particle velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux eq. \ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition eq. \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_5)$,
	showing the non-equivalence of these two definitions.}
	\label{fig:fluxdefs}
\end{figure}
Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
Today, there are two main complementary definitions of the bedload flux being applied in stochastic modeling. Other definitions are discussed comprehensively in \citet{Ballio2014,Ballio2018}.

The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016,Ballio2014}: 
\be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
This definition involves the volumetric concentration $c$ of particles and their velocites at the instant they cross the control surface. These quantities require careful interpretation because bedload particles do not constitute a continuous field \citep{Furbish2012a}. Typically one represents the concentration as a sum of indicator or Dirac delta functions \citep[e.g.][]{}.

@@

One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}. This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$. These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}. This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its reliance on both probabilistic rates and scaling relations.
